---
title: "Rmarkdown_assignment2"
author: "Manon Michon"
date: "10/2/2019"
output: html_document
---
# About the assignment
In 1947 Mr Wiener wrote an article stating that boiling points for alphetic alkanes can be predicted using physicochemical properties (Wiener, 1947). In this assignment molecular descriptors will be used in order to create a partial linear regression (PLS) model and investigate whether it is possible to predict the boiling points with such model.

# Packages

In order to use the functions needed to execute the code, packages need to be installed and added to the library. Since packages only have to be installed once, they are commented out. Please uncomment and run the install.packages lines if the needed packages still need to be installed. Already installed packags don't need to be installed again and can simply be loaded and attached to the library using the 'library' function. The packages loaded here are all necessary for the correct execution of the code.

```{r warning = FALSE, message = FALSE}

#install.packages("WikidataQueryServiceR")
#install.packages("rJava")
#install.packages("rcdk")
#install.packages("pls")

# Load and attach required packages. 
library("WikidataQueryServiceR")
library("rJava")
library("rcdk")
library("pls")

```

#Obtaining the data

Before the model can be constructed, data needs to be obtained. The data used in this model is obtained from wikidata using a SPARQL query and is afterwards tested for assumptions. Physicochemical properties such as molecular descriptors are obtained from the github of the rcdk package.

## The SPARQL query

The function 'query_wikidata' is able to use the written SPARQL query (stored in 'sparql_query') to request the data from wikidata. The result will be stored in a dataframe ('wikidata_query_result). Since we are interested in data with solely 'Kelvin' as unit, the obtained dataset is subset into a dataframe only consisting of data with 'Kelvin' as unit ('subset_query').
Alphatic alkanes can also be requested from wikidata using another unit for the boiling point temperature such as 'Celcius' or 'Fahrenheit'. However, it is recommended to keep using 'Kelvin' since this is the official scientific notation and a lot of alkanes will only have a boiling point registered in 'Kelvin'.

``` {r}
# Defining the SPARQL query
sparql_query <- 'SELECT DISTINCT ?comp ?compLabel ?bp ?bpUnit ?bpUnitLabel ?SMILE WHERE {
    ?comp wdt:P31/wdt:P279* wd:Q41581 ;
  p:P2102 [
    ps:P2102 ?bp ;
    psv:P2102/wikibase:quantityUnit  ?bpUnit
    ] ;
  wdt:P233 ?SMILE .
  SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en". }
}'
wikidata_query_result = query_wikidata(sparql_query); # Requesting the query from the wikidata server
# Distinguishing the data with the correct unit for the boiling point and subsetting that data
subset_query = wikidata_query_result[which(wikidata_query_result$bpUnitLabel == 'kelvin'),]; 
```

## Testing assumptions

The subsetted data is duplicated ('test_assump_subset') in order to edit the data to check it without having the chance of editing the original subsetted dataset by accident. The data is checked by ordering the boiling points from low to high, and then plotting them into a scatterplot. Outliers can be seen when points do not follow the general trend displayed in the scatterplot. The boiling points are plotted on the y-axis. All points seem to follow the general trend, and no outliers are detected. The general trend is however not linear, which requires further testing for normal distribution. From the histogram it can be concluded that the dataset is not normally distributed. A scatterplot from the dataset in the original order shows that the data -however not normally distributed- is sufficiently unordered to be able to randomly select a train and test dataset.

``` {r}
test_assump_subset = subset_query; # Duplication of the original dataset to be sure the original dataset stays intact
# Ordering the dataset from low to high boiling point
test_assump_subset = test_assump_subset[order(test_assump_subset$bp, decreasing = FALSE),] 

# Plotting the boiling points ordered from low to high boiling point to check for outliers
plot(test_assump_subset$bp, main = "Scatterplot of query data", ylab = "Boiling point") 

# Histogram to check for normal distribution
hist(test_assump_subset$bp, main = "Histogram of query data", xlab = "Boiling point") 

# Checking for inequality of data distribution in the original dataset by plotting the data with no order in boiling point
plot(subset_query$bp, main = "Distribution of query data", ylab = "Boiling point") 

```

## Molecular descriptors and dataset editing

Certain columns from the query subset are extracted and renamed to use them at a further point in the process.
The chosen physicochemical properties are retrieved from the rcdk github for the selected parsed SMILES. These results are stored in 'descs' as a dataframe, to which the boilingpoints from the wikidata query are added. From this, two subsets are made: a training-set and a test-set. The splitting of the data into a training- and test-set can be done in any matter, based on the random distribution seen above. The way the data is split does however affect the error calculated further on.

``` {r}
bps_measured = subset_query[,3]; # Renaming column 3 of the subsetted query data to further use this column
smiles = subset_query[,6]; # Renaming column 6 of the subsetted query data to further use this column

mols = parse.smiles(subset_query[,6]); # Parses the SMILES given
# Selection of physicochemical properties requested
descNames <- c( 'org.openscience.cdk.qsar.descriptors.molecular.WienerNumbersDescriptor',
                'org.openscience.cdk.qsar.descriptors.molecular.APolDescriptor',
                'org.openscience.cdk.qsar.descriptors.molecular.CarbonTypesDescriptor',
                'org.openscience.cdk.qsar.descriptors.molecular.AtomCountDescriptor'
                );
descs = eval.desc(mols, descNames); # Request the properties for the selected SMILES

desc_dataset = cbind(descs, bps_measured) # Binding the molecular descriptors with boilingpoints
desc_test = desc_dataset[1:20,]; # Establishing a test dataset
desc_train = desc_dataset[21:114,] # Establishing a train dataset
```


# The Model

## Construction of the Model and training

A Partial Least Squares Regression model is constructed using the PLS package to investigate whether the boiling points of the aliphatic alkanes can be predicted. The model is validated using the Leave-One-Out validation method from the PLS package.
The RMSEP (Root Mean Square Error of Prediction) plot show the expected error for each number of components used in the model. Based on this graph, the number of component used for testing will be 2. The error is more or less the same for all number of components higher than one. Using 2 will ensure the simplest model possible.
The scatterplot of the model shows the distribution of the trainingset data along the regression line. The scatterplot is shown here for 2 components since this is the number of components that will be used for the test-dataset.

``` {r}
PLStest = plsr(bps_measured ~ ., data = desc_train, validation = "LOO"); # Partial Least Square model
plot(RMSEP(PLStest), legendpos = "topright", main = "RMSEP plot") # Plot the error for each amount of latent variables considered
plot(PLStest, ncomp = 2, line = TRUE) # Visualisation of the PLS model
```

## Testing of the model

Now that the model has been constructed, it can be tested. The predict function will use the previously constructed PLS model to try to predict boiling points based on the molecular descriptors of the data in the test dataset. The number of components used here is 2. This was found to be the simplest model while also being within a reasonable error range.
The result was plotted in a scatterplot. The regression line shown here visualises the model.
The RMSEP function displays again the error of the model for each number of components The fact that the error here was the lowest for 2 components confirms that this was choosen correctly.


``` {r}
predict(PLStest, ncomp = 2, newdata = desc_test) # Prediction of boilingpoints using the test data
predplot(PLStest, ncomp = 2, newdata = desc_test, asp = 1, line = TRUE) # The test data is plotted
RMSEP(PLStest, newdata = desc_test) # RMSEP for each of the possible components for the test dataset
```



